Online Social Cohesion Reflects Real-World Group Action in Syria during the Arab Spring

### 2020-12-23 manuscript.Rmd
Update intro pt. 1

abstract: |
  In recent years, political activists have taken to social media platforms to
  rapidly reach broad audiences. Despite the prevalence of micro-blogging in
  these sociopolitical movements, the degree to which virtual mobilization
  reflects or drives real-world movements is unclear. Here, we explore the
  dynamics of real-world events and Twitter social cohesion in Syria during the
  Arab Spring. Using the nonlinear methods cross-recurrence quantification
  analysis and windowed cross-recurrence quantification analysis, we investigate
  if frequency of events of different intensities are coupled with social
  cohesion found in Syrian tweets. Results indicate that virtual social cohesion
  is coupled with the counts of all and negative events each day but shows a
  more nuanced connection to positive events, suggesting that virtual
  mobilization does reflect real-world movements. We conclude with a discussion
  of implications and applications of nonlinear methods in political science
  research.
  <!-- https://tinyurl.com/ybremelq -->

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

In December 2010, a Tunisian man named Mohamed Bouazizi was unjustifiably
stripped of his vendor cart by municipal officials. His humiliation led him to
set himself on fire in protest to the authoritarian rule, leading to his death
several days later. His actions served as a catalyst for the Tunisian
Revolution, and one week after his death, Zine El Abidine Ben Ali---the
longstanding authoritarian president---fled the country. The events of Tunisia
rapidly spread to surrounding countries also subject to unjust authoritarian
rule, including Egypt, Libya, and Syria.

Following a wave of democratization efforts across the Middle East and North
Africa, modest Syrian protests began in March 2011. The regime responded harshly
with imprisonments, detentions, censorship, and military operations as the
resistance to Assad's regime spread. The fighting escalated dramatically in
March 2012, after increased unrest in the western city of Homs. International
actors---including the Red Cross/Red Crescent Society, the United Nations
General Assembly, the U.N. Security Council, and individual world
leaders---worked toward cease-fires, negotiations, and concessions without much
success. While many high-level military and diplomatic officials defected,
numerous loyal forces continued the regime's repression, and the Syrian people
turned to social media to coordinate their activism and communicate to the
world. Between March and June 2012, more than 8 million tweets were sent within
Syria, revealing the details of the conflict suppressed by the government.

@@

## Social Media and Social Movements
Massive online social cohesion can manifest in real-world social behavior.
Social media platforms enable online social cohesion by offering distributed
information-gathering and real-time information dissemination. Given its
accessibility and proficiency as an information exchange tool, Twitter has
provided social and political activists an opportunity to complement or even
replace real-world social mobilization. During the 2012 Italian protests for
global economic change, tweeting was determined to be more effective locally in
discussing real-world, real-time events than traditional mainstream media
platforms [@vicari2013public]. As people increasingly rely on social media for
local news and information, we must understand exactly how these micro-blogs
connect to and reflect real-world action.

Along with Facebook, Twitter has been credited as essential to the Arab
Spring---both for protesters and for the government's response to them
[@comninos2011twitter]. The Arab Spring has been deeply examined within the
political science literature [e.g., @lutterbeck2013arab;
@koehler2017political], but to our knowledge, the direct *coupling* between
Syrian Twitter cohesion and international action has not been investigated. In
fact, some have been skeptical about the effect of social media on collective
action at all [@seay2014slacktivism].

@@

Citizens communicate with each other with increased transparency due to the
ubiquitous availability of social media such as low-cost and egalitarian
platforms like Twitter. Citizens are able to access these tools via mobile
phones and computers with Internet connectivity, providing a stream of real-time
shared information. Citizens can share the location of events like protests and
military incursions, the degree of state repression, and the number of
casualties. In turn, the government---interested in remaining in
power---monitors social media platforms to ascertain emerging threats
[@comninos2011twitter], such as where new groups are forming, where existing
groups are strengthening, where gatherings are held, and who the leaders are.

An example of a leader expressing his interest in such public
conversation is given by Bashar al-Assad in a speech on June 3, 2012:

\noindent I know that I have been away from the media for a long time, and I
have missed having direct contact with the citizens, but I have always been
following up with the daily occurrences and gathering the information so that my
speech can be built on what is said by the street.

\noindent With few exceptions, Assad chose to keep social media platforms up and
running. Gohdes [-@gohdes2015pulling] notes that regimes can attempt to thwart
the growth of social protest online and in the streets by shutting down the
Internet. During these times, military repression increases, but government
information about the citizens decreases. As a result, regimes may elect to
maintain Internet connectivity to monitor and surveil their citizens.

At the same time, the insulation of online social platforms allows researchers
to detail real-world social dynamics at unprecedented resolution and scale
[@paxton2017finding]. The digital traces left by individuals and collectively as
groups are regarded as a valuable source of information that can offer powerful
insights into states of social disequilibrium, uprisings, and escalations into
civil wars. By examining these conversations, researchers can study the degree
of cohesion between speakers and the alignment between that cohesion and their
collective political behavior.

@@

The type of collective contentious behavior that manifests in society is a
function of the available political opportunity structures. The political
opportunity structure is “comprised of *(sic)* specific configurations of
resources, institutional arrangements and historical precedents for social
mobilization, which facilitate the development of protest movements in some
instances and constrain them in others” [@kitschelt1986political]. What emerges
as strategies and tactics in contentious politics are a function of
what the system makes possible. This includes the number of people and their
disposition (i.e. nonviolent or violent), as well as their internal structure.
The more cohesive a group is, the greater threat they pose to the establishment.

>>>>

abstract: |
  In recent years, political activists have taken to social media platforms to rapidly reach broad audiences. Despite the prevalence of micro-blogging in these sociopolitical movements, the degree to which virtual mobilization reflects or drives real-world movements is unclear. Here, we explore the dynamics of real-world events and Twitter social cohesion in Syria during the Arab Spring. Using the nonlinear methods cross-recurrence quantification analysis and windowed cross-recurrence quantification analysis, we investigate if frequency of events of different intensities are coupled with social cohesion found in Syrian tweets. Results indicate that online social cohesion is coupled with the counts of all and negative events each day but shows a more nuanced connection to positive events, suggesting that online mobilization does reflect real-world movements. We conclude with a discussion of implications and applications of nonlinear methods in political science research.
  <!-- https://tinyurl.com/ybremelq -->

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
#output            : papaja::apa6_pdf
---

@@

In December 2010, a Tunisian man named Mohamed Bouazizi died after setting himself on fire to protest the authoritarian government’s unjustifiable confiscation of his vendor cart by municipal officials. His actions served as a catalyst for the Tunisian Revolution, which rapidly spread to surrounding countries also subject to unjust authoritarian rule, including Egypt, Libya, and Syria.

Following a wave of democratization efforts across the Middle East and North
Africa, modest Syrian protests began in March 2011. The regime responded harshly
with imprisonments, detentions, censorship, and military operations as the
resistance to Assad's regime spread. The fighting escalated dramatically in
March 2012, after increased unrest in the western city of Homs. International
actors---including the Red Cross/Red Crescent Society, the United Nations
General Assembly, the U.N. Security Council, and individual world
leaders---worked toward cease-fires, negotiations, and concessions without much
success. While many high-level military and diplomatic officials defected,
numerous loyal forces continued the regime's repression, and the Syrian people
turned to social media to coordinate their activism and communicate to the
world. Between March and June 2012, millions of tweets were sent within
Syria, revealing the details of the conflict suppressed by the government. A full account of the history of the Arab Spring is outside of the scope of the current article; more information can be found in Brownlee (2013) [fix bib] and Lynch (2016) [fix bib].

@@

## Social Media and Social Movements
Massive online social cohesion can manifest in real-world social behavior.
Social media platforms enable online social cohesion by offering distributed
information-gathering and real-time information dissemination. Given its
accessibility and proficiency as an information exchange tool, Twitter has
provided social and political activists an opportunity to complement (Greijdanus et al., 2020) [fix bib] and not hinder participation in (Harlow & Harp, 2012; Milošević-Đorđević & Žeželj, 2017) [fix bib] real-world social mobilization. During the 2012 Italian protests for global economic change, tweeting was determined to be more effective locally in discussing real-world, real-time events than traditional mainstream media platforms [@vicari2013public]. As people increasingly rely on social media for local news and information, we must understand exactly how these micro-blogs connect to and reflect real-world action.

Along with Facebook, Twitter has been credited as essential to the Arab
Spring---both for protesters and for the government's response to them
[@comninos2011twitter]. The Arab Spring has been deeply examined within the
political science literature [e.g., @lutterbeck2013arab;
@koehler2017political], but to our knowledge, the direct *coupling* between
Syrian Twitter cohesion and international action has not been investigated  (though hashtag usage has been linked to next-day protests; Steinert-Threlkeld et al., 2015) [fix bib]. In fact, some have been skeptical about the effect of social media on collective action at all [@seay2014slacktivism].

@@

Protestors and social activists communicate with each other with increased transparency due to the ubiquitous availability of social media like Twitter. Activists are able to access these tools via mobile
phones and computers with Internet connectivity, providing a stream of real-time
shared information. Citizens can share the location of events like protests and
military incursions, the degree of state repression, and the number of
casualties. In turn, the government---interested in remaining in
power---monitors social media platforms to ascertain emerging threats
[@comninos2011twitter], such as where new groups are forming, where existing
groups are strengthening, where gatherings are held, and who the leaders are.

With few exceptions, leader Bashar al-Assad chose to keep social media platforms up and running. Gohdes [-@gohdes2015pulling] notes that regimes can attempt to thwart the growth of social protest online and in the streets by shutting down the Internet. During these times, military repression increases, but government information about the citizens decreases. As a result, regimes may elect to maintain Internet connectivity to monitor and surveil their citizens.

At the same time, the insulation of online social platforms allows researchers
to detail real-world social dynamics at unprecedented resolution and scale
[@paxton2017finding]. The digital traces left by individuals and collectively as
groups are regarded as a valuable source of information that can offer powerful
insights into states of social disequilibrium, uprisings, and escalations into
civil wars. By examining these conversations, researchers can study the degree
of cohesion between speakers and the alignment between that cohesion and their
collective political behavior, just as the degree of cohesion among real-world protestors in past movements has been captured in their shared sign messaging (Windsor, 2020) [fix bib].

@@

The type of collective contentious behavior that manifests in society is a
function of the available political opportunity structures. The political
opportunity structure is “comprised of *(sic)* specific configurations of
resources, institutional arrangements and historical precedents for social
mobilization, which facilitate the development of protest movements in some
instances and constrain them in others” [@kitschelt1986political]. Emergent strategies and tactics in contentious politics are a function of
what the system makes possible. This includes the number of people and their
disposition (i.e. nonviolent or violent), as well as their internal structure.
The more cohesive a group is, the greater threat they pose to the establishment.
